---
title: "Replication of WGCNA Modules in Arijs 2018 data"
output: html_document
editor_options: 
chunk_output_type: console
---

# Load in the replication data

```{r echo=FALSE}
#load BioFunk
source("~/devel/Useful_Scripts/BioFunk.R")
library(here)
#load WGCNA
suppressPackageStartupMessages(library(WGCNA))
allowWGCNAThreads()
options(stringsAsFactors = FALSE)
suppressPackageStartupMessages(library("GEOquery"))
suppressPackageStartupMessages(library("affy"))
library(knitr)
suppressPackageStartupMessages(library(ggbeeswarm))
suppressPackageStartupMessages(library(ggpubr))
set.seed(666)

theme_set(theme_cowplot())

#### commands used to generate the normalised data ####
#getGEOSuppFiles("GSE73661")
#untar("GSE73661_RAW.tar",exdir="data")
#cels=list.files("data/",pattern="CEL")
#setwd("data/")
#raw.data = ReadAffy(verbose = FALSE, filenames = cels)
#data.rma.norm = rma(raw.data)
#rma = exprs(data.rma.norm)
#write.table(rma, file = "normalised_array_exp.tsv", quote = FALSE, sep = "\t")

#read in the normalised data
#expdat=read.table("normalised_array_exp.tsv",header=T,sep="\t")
#mapped=t(expdat)
#convert the probes names to gene ids
#library(hugene10sttranscriptcluster.db)
#mapping=data.frame(hugene10sttranscriptclusterENSEMBL)
#colnames(mapped)=mapping$ensembl_id[match(colnames(mapped),mapping$probe_id)]
#mapped=mapped[,-which(is.na(colnames(mapped)))]
#mapped=mergDupCol(mapped,TRUE)
#rownames(mapped)=gsub(".CEL.gz","",rownames(mapped))
#clean_expr_dat=t(mapped)
#write.table(clean_expr_dat,"cleaned_expr_dat.tsv",sep="\t")

repdat=read.table(here("Data","Input","Arijs_2018_cleaned_expr_dat.tsv"), sep="\t")
colnames(repdat)=lapply(strsplit(colnames(repdat),"_"),"[[",1)

#### Original code to generate the metadata table ####

#load the metadata downloaded from GEO
#geodat=getGEO(filename="GSE73661_series_matrix.txt.gz")
#geometa=pData(geodat)
#metasmall=geometa[,c(1,2,36,37,38,39,40)]
#write.table(metasmall,"cleaned_meta_table.tsv",sep="\t")

metadat=read.table(here("Data","Input","Arijs_2018_cleaned_meta_table.tsv"),sep="\t")
metadat=data.frame(lapply(metadat,factor))
rownames(metadat)=metadat$geo_accession
metadat=metadat[match(rownames(metadat),colnames(repdat)),]

repdat=repdat[,match(colnames(repdat),rownames(metadat))]

```

Microarray data was downloaded from the Arijs et al. 2018 Gut paper as raw image data from (GEO:GSE73661).
Post-processing, this contained **`r nrow(metadat)`** samples and **`r nrow(repdat)`** genes.
Microarray data was normalised using the robust multi-array average (RMA) measure.

# Map to the modules from the discovery dataset

```{r, echo=FALSE}
#our module data
moddefs=read.csv(here("Data","Generated","discovery_WGCNAgeneInfo.csv"))
modcors=read.csv(here("Data","Generated","discovery_WGCNAmodInfo.csv"))
moddefs=moddefs[-which(moddefs$moduleColor=="unassigned"),]

#list of modules significantly associated with NANCY score
signancy=modcors[modcors$NANCYscore_Pvalue_FDR_Pvalue<0.05,]

#match replication data and module defs
matchdef=moddefs[moddefs$geneid%in%rownames(repdat),]
matchrep=repdat[match(matchdef$geneid,rownames(repdat)),]

#calculate the module coverages in the replication data set (gene % observed)
modcov=data.frame(Original=table(moddefs$moduleColor),Replication=table(matchdef$moduleColor))
modcov=modcov[,-3]
colnames(modcov)=c("Module","Original Size (No. Genes)","Replication Size (No. Genes)")
modcov$Percentage=(modcov[,3]/modcov[,2])*100
modcov$NANCYsig=ifelse(modcov$Module%in%signancy$X,1,0)
colnames(modcov)[ncol(modcov)]="FDR Sig. with NANCY Score in discovery cohort."
modcov=modcov[order(modcov$Percentage,decreasing = T),]

```

In our unbiased analysis we had **`r nrow(moddefs)`** genes assigned to a module (excluding the genes assigned to grey).
From these, **`r nrow(matchdef)`** were also in the replication data set (**`r round((nrow(matchdef)/nrow(moddefs))*100)`%**).

## Module overlaps

`r kable(modcov,format="markdown",row.names=F)`

## Prepare data

Responders were defined as anyone who had an endoscopic score of 0 or 1 at any of the time points during the study.

```{r data prep,include=F}
#generate the module eigengenes
RepMEList = moduleEigengenes(t(matchrep), colors = matchdef$moduleColor[,drop=TRUE], excludeGrey = TRUE)
RepMEs = RepMEList$eigengenes
colnames(RepMEs)=gsub("ME","",colnames(RepMEs))

#remove non NANCY associated MEs
RepMEs = RepMEs[,colnames(RepMEs)%in%signancy$X]
#write out the pretreatment MEs
write.csv(RepMEs[metadat$week..w..ch1=="W0",],here("Data","Generated","Arijs_2018_pre_treatment_MEs.csv"))

#Note responder at any time point
repmet=metadat[metadat$week..w..ch1=="W0",]
#remove placebo
repmet=repmet[-which(repmet$induction.therapy_maintenance.therapy.ch1=="plac_plac"),]
repmet$study.individual.number.ch1=factor(repmet$study.individual.number.ch1)
repmet=repmet[,c(3,4,5)]
#keep note of who is not on vdz
repmet$vdz_comb=ifelse(repmet$induction.therapy_maintenance.therapy.ch1=="IFX",0,1)
response=c()
for(i in repmet$study.individual.number.ch1){
  samps=metadat[metadat$study.individual.number.ch1==i,]
  if(0%in%samps$mayo.endoscopic.subscore.ch1||1%in%samps$mayo.endoscopic.subscore.ch1){
    response=c(response,1)
  }
  else{
    response=c(response,0)
  }
}
repmet$treatment_response=response
repmet=repmet[,-c(3)]

#write out response data for use elsewhere
write.csv(repmet,here("Data","Generated","Arijs_2018_w0_with_repsonse_meta.csv"))

#subset data to just pre-treatment timepoints for comparisons
RepMEs=RepMEs[rownames(metadat)%in%rownames(repmet),]

```

## Pre-treatment expression vs response - split by regimen 

This study had a complex set-up whereby patients were treat on different regimens of anti-integrin therapy with an additional group treated with anti-TNF. To determine module levels in responders/non-responders in these groups were perform ANOVA to indentify modules different between responders by any treatment regime that carry out post-hoc to tests to identify within which treatment regimens that module is different between responsers and non-responders.

```{r, fig.height=5, fig.width=12,results="asis",echo=F}


plots=list()
plotspair=list()
ps=c()

#get the color mapping from the name
namekey=read.csv(here("Data","Input","module_number_key.csv"),header=F)

for(i in 1:ncol(RepMEs)){
  me=RepMEs[,i]
  mename=colnames(RepMEs)[i]
  groups=factor(repmet$treatment_response)
  regime=repmet$induction.therapy_maintenance.therapy.ch1
  colorname=namekey[namekey[,2]==mename,1]
  df=data.frame("Eigengene"=me,"Response"=groups,"Regimen"=regime)
  pl=ggplot(df,aes(y=Eigengene,x=Regimen,fill=Response))+geom_boxplot()+geom_quasirandom()+
    ylab(paste0(mename))+ggtitle(paste0(mename))+scale_fill_manual(values=c(colorname,"grey"))
  plots[[mename]]=pl
  plp=ggplot(df,aes(y=Eigengene,x=Regimen,fill=Response))+geom_boxplot()+geom_quasirandom()+
    ylab(paste0(mename))+ggtitle(paste0(mename))+scale_fill_manual(values=c(colorname,"grey"))+stat_compare_means(method="wilcox")
  plotspair[[mename]]=plp
  test=anova(lm(me~groups+regime))
  ps=c(ps,test$`Pr(>F)`[1])
}

adjp=p.adjust(ps,method="fdr")
siglist=plots[adjp<0.05]

cat("**ANOVA results responders vs. non-responders split by (and adjusting for) different treatments**")
psdf=data.frame("Module"=colnames(RepMEs),"ANOVA_p"=ps,"ANOVA_adj_p"=adjp)
psdf=psdf[order(adjp),]
kable(psdf)

write.csv(psdf,here("Data","Generated","Arijs_2018_split_by_treatment_anova_eigengenes.csv"))

cat("**Boxplots for FDR significant results**")
plot_grid(plotlist = siglist[c(psdf$Module[psdf$ANOVA_adj_p<0.05])], ncol=2)

cat("**Plots to explore M5 and M4**")
plot_grid(plotlist=plots[c("M5","M4")], ncol=2)
ggsave(here("Plots","Arijs_M5_M4_boxplots.pdf"),width=20,height=8)

cat("**Pairwise test for M5 only for paper**")
plot(plotspair$M5)
ggsave(here("Plots","Arijs_M5_pairwise.pdf"),width=12,height =8)

```