---
title: "Discovery Cohort WGCNA Analysis"
editor_options: null
output:
  html_document: default
  pdf_document: default
chunk_output_type: console
---

# Data QC

```{r echo=FALSE}
#TOGGLE THIS TO TURN ON NETWORK PLOTS - These require the TOM to be built which is a large file so toggle off if the TOM has been removed to save space
netplots=FALSE

#library here lets things be relative to this repo
library(here)
#load BioFunk helper script
source(here("Scripts","BioFunk.R"))

#load other libraries
library(pheatmap)
library(biomaRt,quietly = T)
library(knitr,quietly = T)
#load WGCNA
suppressPackageStartupMessages(library(WGCNA))
options(stringsAsFactors = FALSE)
allowWGCNAThreads(nThreads = 20)
#set seed to make reproducible
set.seed(666)

theme_set(theme_cowplot())

#load in tpm counts
tpmcounts=read.csv(here("Data","Input","discovery_tpm_counts.csv.gz"))

#set rownames to gene_id and remove non-count columns (columns 4 onwards)
tpmformat=tpmcounts[,4:ncol(tpmcounts)]
rownames(tpmformat)=tpmcounts$gene_id

#rename samples to numbers
sampnums=unlist(lapply(strsplit(colnames(tpmformat),"_"),"[[",5))
colnames(tpmformat)=sampnums

#filter so that genes are non-zero in at least 50% of samples
tpmfilt=zeroFilt(tpmformat,0.5)

#log and transpose
tpm=t(log(tpmfilt+1))

```

## Genes

TPM expression data loaded contains **`r ncol(tpmcounts)`** samples and **`r nrow(tpmcounts)`** genes.
This was filtered to only retain genes observed in at least 50% of samples, resulting in **`r ncol(tpm)`** genes.
TPM counts were then log transformed *(log(x+1))* for further analyses.

```{r echo=F}
#remove samples ommited from DE analysis due to failing QC
samples2remove=read.csv(here("Data","Input","failed_samples.csv.gz"))
tpm=tpm[!rownames(tpm)%in%samples2remove[,1],]

```

## Samples

Removed samples flagged for omission after failing QC in DE analysis, data now contains **`r nrow(tpm)`** samples.
Heirarchical clustering plot to check for outliers and metadata associations.

```{r,fig.width=10,fig.height=4,echo=FALSE}
traitData = read.csv(here("Data","Input","discovery_metadata.csv.gz"),stringsAsFactors = T)
#subset metadata to remaning samples and put in matching order
traitData=traitData[match(rownames(tpm),traitData$sample_ID..patient.),]

#reformat disease data
traitData$Disease.CRC=as.factor(ifelse(traitData$disease=="A",1,0))
traitData$Disease.UC=as.factor(ifelse(traitData$disease=="B",1,0))
traitData$Disease.CD=as.factor(ifelse(traitData$disease=="C",1,0))
traitData$Combined=as.factor(ifelse(traitData$disease=="B"|traitData$disease=="C",1,0))
#remove unused columns
traitData=data.frame(lapply(traitData,as.character))
traitData=traitData[,-c(1,3,4)]
traitData[traitData=="A"]=0
traitData[traitData=="B"]=1
traitData[traitData=="Y"]=1
traitData=data.frame(lapply(traitData,as.factor))
rownames(traitData)=rownames(tpm)
continuous=c(1,3,6,8)
for(i in continuous){
  traitData[,i]=as.numeric(as.character(traitData[,i]))
}

#plot a dendrogram with the metadata as color bars
metaDend(tpm,traitData,10,1/3,102,4)

```

No obvious outliers, reasonable clustering of control and IBD samples.

# WGCNA

## Define co-expression modules

Find a soft-threshold power to define a network between the genes. Using signed weights to define the network.

```{r, warning=FALSE,fig.width=10,cache=TRUE,echo=FALSE}
# Choose a set of soft-thresholding powers
powers = c(c(1:10), seq(from = 12, to=20, by=2))
# Call the network topology analysis function, uses Pearson correlation as default
sft = pickSoftThreshold(tpm, powerVector = powers, verbose = 5, networkType = "signed")
# Plot the results:
par(mfrow = c(1,2));
# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
main = paste("Scale independence"))
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
labels=powers,cex=0.9,col="red")
# this line corresponds to using an R^2 cut-off of h (we want our fit to be above this)
abline(h=0.8,col="red")
# Mean network connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,5],
xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
main = paste("Mean connectivity"))
text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=0.9,col="red")
```

Use a soft-thresholding power of **9** to generate an adjaceny matrix of the network this threshold produces.
Then generate a topological overlap matrix from the adjacency matrix using a block-wise approach.

```{r echo=FALSE }
if(file.exists(here("Data","Generated","bwnet.Rds"))){
  bwnet=readRDS(here("Data","Generated","bwnet.Rds"))
}else{
#write out the filtered TPM table
saveRDS(tpm,here("Data","Generated","discovery_filtered_tpm.Rds"))
#use the blockwise method to handle the large number of genes
bwnet = blockwiseModules(tpm, maxBlockSize = 30000, power = 9, TOMType = "signed", minModuleSize = 30,
reassignThreshold = 0, mergeCutHeight = 0.1, numericLabels = TRUE, saveTOMs = TRUE,
saveTOMFileBase = "Data/Generated/TOM", verbose = 3)
#save the bwnet object as it is resource heavy to make
saveRDS(bwnet,here("Data","Generated","bwnet.Rds"))
}
# Plot the dendrograms
bwModuleColors = labels2colors(bwnet$colors)
par(mfrow = c(length(bwnet$dendrograms),1))
pdend=function(net,block,ncols){
plotDendroAndColors(net$dendrograms[[block]], ncols[net$blockGenes[[block]]],
"Module colors", main = paste0("Gene dendrogram and module colors in block ",block),
dendroLabels = FALSE, hang = 0.03,
addGuide = TRUE, guideHang = 0.05)}
for(i in c(1:length(bwnet$dendrograms))){
pdend(bwnet,i,bwModuleColors)}

```

Look at correlations between the module eigengenes.

```{r fig.height=5,fig.width=10,echo=FALSE}
# Calculate eigengenes
MEList = moduleEigengenes(tpm, colors = bwModuleColors, excludeGrey = TRUE,)
MEs = MEList$eigengenes

#correlation between the modules
# Calculate dissimilarity of module eigengenes
MEDiss = 1-cor(MEs);
# Cluster module eigengenes
METree = hclust(as.dist(MEDiss), method = "average");
# Plot the result
ggdendrogram(METree,rotate = F)

```

After filtering a total of **`r length(which(bwnet$goodGenes==TRUE))`** genes were included in module identification.
These were clustered into modules by splitting data into **`r length(bwnet$dendrograms)`** blocks.
A total of **`r length(table(bwnet$colors))-1`** modules were identified (plus an additional unassigned module (grey)).

## Module correlations with metadata

Find correlation between module eigengenes and metadata traits.
Find modules associated with inflammation.

```{r, fig.width=15,fig.height=8, results='asis',echo=FALSE}
#get number of samples
nSamples=nrow(tpm)
# Reorder MEs so that correlated MEs are side by side
MEs = orderMEs(MEs)

#write out the MEs
rownames(MEs)=rownames(traitData)
write.csv(MEs,here("Data","Generated","discovery_MEs.csv"))

#rename the MEs from their color names to the numbers used int he manuscript
namekey=read.csv(here("Data","Input","module_number_key.csv.gz"),header=F)
colnames(MEs)=gsub("ME","",colnames(MEs))
colnames(MEs)=namekey[match(colnames(MEs),namekey[,1]),2]

#correlate MEs vs numerical versions of traits
numTraitData=data.frame(lapply(traitData,as.numeric))
moduleTraitCor = cor(MEs, numTraitData, use = "p")
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nSamples)
padj=matrix(p.adjust(moduleTraitPvalue,method="fdr"),ncol=17,byrow = F)
colnames(padj)=colnames(traitData)
rownames(padj)=rownames(moduleTraitPvalue)
#remove traits and modules with no significant associations
nonsigrows=which(rowSums(moduleTraitPvalue<0.05)==0)
nonsigcols=which(colSums(moduleTraitPvalue<0.05)==0)

colorer=colorRampPalette(rev(brewer.pal(11,"RdBu")))

#adjust p significant
nonsigarows=which(rowSums(padj<0.05)==0)
nonsigacols=which(colSums(padj<0.05)==0)
filtadjCor=moduleTraitCor[-nonsigarows,-nonsigacols]
filtadjP=padj[-nonsigarows,-nonsigacols]
filtadjP=ifelse(filtadjP<0.05,1,0)
cat("**Modules/traits with at least one FDR significant correlation (FDR p<0.05)**")
simpHeat(t(filtadjCor),t(filtadjP),colorer(11))

#second cleaned-up version for paper
removevars=c(11,6,10,9,8)
filtadjCorClean2=filtadjCor[,-removevars]
filtadjPClean2=filtadjP[,-removevars]
rownames(filtadjCorClean2)=gsub("ME","",rownames(filtadjCorClean2))
rownames(filtadjPClean2)=rownames(filtadjCorClean2)

simpHeat(t(filtadjCorClean2),t(filtadjPClean2),colorer(11))+scale_fill_gradientn(colors=colorer(11),breaks=c(-0.8,0,0.8),labels=c(-0.8,0,0.8),limits=c(-0.8,0.8))+scale_color_manual(values=c("black",NA),labels=c("FDR<0.05",""),name="Significance")
ggsave(here("Plots","discovery_cleaned_metadata_heatmap_selected_variables.pdf"),,width = 15,height = 8)

pas=data.frame(padj)
ps=data.frame(moduleTraitPvalue)
cs=data.frame(moduleTraitCor)
NANCY=data.frame(Module=rownames(cs),Cor=cs$NANCYscore,p=ps$NANCYscore,FDR=pas$NANCYscore)
#write out list of modules  FDRsignificant with NANCY
write.table(NANCY,here("Data","Generated","discovery_NANCY_module_correlations.tsv"),sep="\t",row.names = F)

```

It is clear that the modules are largely correlating with differences in inflammation status. Especially given the FDR adjusted p-values. Nicely, the medication traits are showing opposing effects. The only traits with no nominally significant module associations were **`r names(nonsigcols)`**. The only modules with no significant associations were **`r names(nonsigrows)`**.

```{r echo=FALSE}
# Write Module Stats to Files
# Define variable weight containing the NANCY score
weight = as.data.frame(traitData$NANCYscore)
names(weight) = "weight"
# names (colors) of the modules
modNames = names(MEs)
#correlation between gene and modules, cor and p-value
geneModuleMembership = as.data.frame(cor(tpm, MEs, use = "p"))
MMPvalue = as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples));
names(geneModuleMembership) = paste("MM", modNames, sep="");
names(MMPvalue) = paste("p.MM", modNames, sep="");
#correlation between gene and inflammed status
geneTraitSignificance = as.data.frame(cor(tpm, weight, use = "p"));
GSPvalue = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificance), nSamples));
names(geneTraitSignificance) = paste("GS_NANCYscore.", names(weight), sep="");
names(GSPvalue) = paste("p.GS_NANCYscore.", names(weight), sep="");

#rename the gene identifiers to gene names using the original input table
geneids=colnames(tpm)
geneids2names=match(geneids,tpmcounts$gene_id)
# Create the starting data frame
geneInfo0 = data.frame(geneid = geneids,
geneName = tpmcounts$gene_name[geneids2names],
moduleColor = bwModuleColors,
geneTraitSignificance,
GSPvalue)
# Order modules by their significance for weight
modOrder = order(-abs(cor(MEs, weight, use = "p")));
# Add module membership information in the chosen order
for (mod in 1:ncol(geneModuleMembership))
{
oldNames = names(geneInfo0)
geneInfo0 = data.frame(geneInfo0, geneModuleMembership[, modOrder[mod]],
MMPvalue[, modOrder[mod]]);
names(geneInfo0) = c(oldNames, paste("MM.", modNames[modOrder[mod]], sep=""),
paste("p.MM.", modNames[modOrder[mod]], sep=""))
}
# Order the genes in the geneInfo variable first by module color, then by geneTraitSignificance
geneOrder = order(geneInfo0$moduleColor, -abs(geneInfo0$GS_NANCYscore.weight));
geneInfo = geneInfo0[geneOrder, ]

#rename the colors to module numbers
geneInfo$moduleColor=namekey[match(geneInfo$moduleColor,namekey[,1]),2]
write.csv(geneInfo, file = here("Data","Generated","discovery_WGCNAgeneInfo.csv"), row.names=F)

#write a table of the module correlations with clinical traits
colnames(moduleTraitCor)=paste0(colnames(moduleTraitCor),"_Cor")
colnames(moduleTraitPvalue)=paste0(colnames(moduleTraitPvalue),"_Pvalue")
padjcopy=padj
colnames(padjcopy)=paste0(colnames(moduleTraitPvalue),"_FDR_Pvalue")
modcorframe=cbind(moduleTraitCor,moduleTraitPvalue,padjcopy)
modcorframe=modcorframe[order(rownames(modcorframe)),order(colnames(modcorframe))]
write.csv(modcorframe,file=here("Data","Generated","discovery_WGCNAmodInfo.csv"))
```

## **GO Term enrichment analysis of modules**

```{r warnings=FALSE,cache=TRUE, echo=FALSE}
#use the biomart library to convert our ensembl to entrez ids
ensembl=useMart("ensembl",dataset="hsapiens_gene_ensembl")
map = getBM(c("ensembl_gene_id","entrezgene_id"), filters=c("ensembl_gene_id","with_entrezgene"), values=list(colnames(tpm),TRUE), mart=ensembl)
#filter ensembl genes with no equivalent entrez ID
keep=which(colnames(tpm)%in%map$ensembl_gene_id)
keepmods=bwModuleColors[keep]
keepgenes=colnames(tpm)[keep]

#convert gene names to entrez
id2entrez=match(keepgenes,map$ensembl_gene_id)
entrez=map$entrezgene[id2entrez]
#carry out GSEA, keep top 10 GO terms enriched in each module
GOenr = GOenrichmentAnalysis(keepmods, entrez, organism = "human", nBestP = 10, verbose = F)
#enrichment terms for each module
tab = GOenr$bestPTerms[[4]]$enrichment
tab$module=namekey[match(tab$module,namekey[,1]),2]
write.table(tab, file = here("Data","Generated","discovery_WGCNAGOEnrichmentTable.csv"), sep = ",", quote = TRUE, row.names = TRUE)

```

A total of **`r length(keepgenes)`** were mapped to Entrez gene IDs and used for the GO enrichment analysis. This was performed across **`r length(unique(keepmods))`** keeping the top 10 enriched terms. Results for some modules of interest are shown below.

```{r,results='asis', echo=FALSE}
#as an example print the results for the modules most strikingly correlated with inflammation
plottab=function(col,tab){
  return(kable(tab[which(tab$module==col),c(1,2,5,6,11,13)]))
}

fpdat=data.frame(padj)
cdat=data.frame(moduleTraitCor)

cat("\n\n### Modules postively correlated with NANCY score (FDR p<0.05)")

for(i in 1:nrow(fpdat)){
  if(fpdat$NANCYscore[i]<0.05&cdat$NANCYscore_Cor[i]>0){
    print(plottab(rownames(fpdat)[i],tab))
  }
}

cat("\n\n### Modules negatively correlated with NANCY score (FDR p<0.05)")

for(i in 1:nrow(fpdat)){
  if(fpdat$NANCYscore[i]<0.05&cdat$NANCYscore_Cor[i]<0){
    print(plottab(rownames(fpdat)[i],tab))
  }
}

```

```{r echo=F}
if(netplots==TRUE){
  source(here("Scripts","plot_module_networks.R"))
}

```


## **Summary of output files**

*discovery_module_analysis.html* - This report showing all the steps in the analysis.

*discovery_WGCNAgeneInfo.csv* - Table showing the module assignment for each gene, its correlation with inflammation, and its correlation with all other modules.

*discovery_WGCNAmodInfo.csv* - Table showing the module correlations with all clinical traits including inflammation.

*discovery_WGCNAGOEnrcihmentTable.csv* - Table contianing the top 10 GO terms enriched for in each module.

*TOM-block.-.Rds* - Topological overlap matrices used to generate the modules (generated blockwise).

*discovery_filtered_tpm.Rds* - The TPMs after filtering used to make the modules. Passed to the network script to name the genes.

*discovery_NANCY_module_correlations.tsv* - Eigengene correlation values and significance for each module vs NANCY scores.

Additional plots not included in the markdown are network plots of the gene networks within each module, which are in Plots/Networks at various TOM thresholds.




