---
title: "Replication of WGCNA Modulesin Haberman data"
output: html_document
editor_options: 
chunk_output_type: console
---

# Load in the replication data

```{r echo=FALSE}
library(here)
#load BioFunk
source("~/devel/Useful_Scripts/BioFunk.R")
suppressPackageStartupMessages(library(hablar))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggbeeswarm))
suppressPackageStartupMessages(library(ggpubr))
library(knitr)
#load WGCNA
suppressPackageStartupMessages(library(WGCNA))
allowWGCNAThreads()
options(stringsAsFactors = FALSE)
suppressPackageStartupMessages(library("GEOquery"))
suppressPackageStartupMessages(library("affy"))
set.seed(666)

theme_set(theme_cowplot())

#### Code used to generate the expression file ####

#ID is GSE109142
#read in the raw data, this table was merged from the per-sample files on the GEO website
#repdat=read.table("merged_GSE109142.txt",header=T,sep="\t",row.names=1)

#read in the metadata, downloaded from GEO
#metadat=data.frame(getGEO(filename="GSE109142_series_matrix.txt",getGPL = F))
#reformat names and make same order as gene exp data
#metadat$sampnames=unlist(lapply(strsplit(as.character(metadat$title)," "),"[[",1))
#metadat=metadat[match(colnames(repdat),rownames(metadat)),]
#sampnames=rownames(metadat)

#drop useless columns
#metadat=metadat[,c(1,48:59)]
##metadat[metadat=="unknown"]=NA
#metadat[metadat=="NA"]=NA
#metadat = metadat %>% retype() %>% data.frame()
#fs=c(4,5,7,9,10,12,13)
#for(i in fs){
# metadat[,i]=factor(metadat[,i])
#}
#rownames(metadat)=sampnames
#remove controls
#rem=which(metadat$diagnosis.ch1=="Control")
#metadat=metadat[-rem,]
#repdat=repdat[,-rem]

#write.csv(repdat,"Haberman_no_controls_gene_exp_tpm.csv")
#write.csv(metadat,"Haberman_no_controls_cleaned_meta.csv")

repdat=read.csv(here("Data","Input","Haberman_no_controls_gene_exp_tpm.csv.gz"),row.names = 1)
metadat=read.csv(here("Data","Input","Haberman_no_controls_cleaned_meta.csv.gz"),row.names = 1)

```

RNASeq data was downloaded as pre-processed TPMs (from Kalisto) for the Haberman et al. 2019 Nature Comms. paper (GEO:GSE109142).
This contained **`r nrow(metadat)`** samples from pediatric UC patients and included **`r nrow(repdat)`** genes (some filtering in GEO upload).

# Map to modules from discovery data

```{r, echo=FALSE}
#our module data
moddefs=read.csv(here("Data","Generated","discovery_WGCNAgeneInfo.csv"))
modcors=read.csv(here("Data","Generated","discovery_WGCNAmodInfo.csv"))
moddefs=moddefs[-which(moddefs$moduleColor=="unassigned"),]

#list of modules significantly associated with NANCY score
signancy=modcors[modcors$NANCYscore_Pvalue_FDR_Pvalue<0.05,]

#match replication data and module defs
matchdef=moddefs[moddefs$geneName%in%rownames(repdat),]
matchrep=repdat[match(matchdef$geneName,rownames(repdat)),]

#calculate the module coverages in the replication data set (gene % observed)
modcov=data.frame(Original=table(moddefs$moduleColor),Replication=table(matchdef$moduleColor))
modcov=modcov[,-3]
colnames(modcov)=c("Module","Original Size (No. Genes)","Replication Size (No. Genes)")
modcov$Percentage=(modcov[,3]/modcov[,2])*100
modcov$NANCYsig=ifelse(modcov$Module%in%signancy$X,1,0)
colnames(modcov)[ncol(modcov)]="FDR Sig. with NANCY Score in discovery cohort."
modcov=modcov[order(modcov$Percentage,decreasing = T),]

```

In our unbiased analysis we had **`r nrow(moddefs)`** genes assigned to a module (excluding the genes assigned to grey).
From these, **`r nrow(matchdef)`** were also in the replication data set (**`r round((nrow(matchdef)/nrow(moddefs))*100)`%**).

## Module overlaps

`r kable(modcov,format="markdown",row.names=F)`

# Module eigengene comparisons pre-treatment between responders and non-responders

We extracted patients receiving corticosteroid therapy (via either IV or oral administration) from the Haberman study and compared the module eigengene values pre-treatment between the cases and controls. 

```{r, fig.height=15, fig.width=10,results="asis",echo=F}
#generate the module eigengenes
RepMEList = moduleEigengenes(t(matchrep), colors = matchdef$moduleColor, excludeGrey = TRUE,)
RepMEs = RepMEList$eigengenes
colnames(RepMEs)=gsub("ME","",colnames(RepMEs))

#remove non NANCY associated MEs
RepMEs = RepMEs[,colnames(RepMEs)%in%signancy$X]

csMEs=RepMEs[which(metadat$initial.treatment.ch1!="5ASA"),]
csMeta=metadat[which(metadat$initial.treatment.ch1!="5ASA"),]

colnames(csMeta)=gsub(".ch1","",colnames(csMeta))

#get the color mapping from the name
namekey=read.csv(here("Data","Input","module_number_key.csv.gz"),header=F)

csplots=list()
ps=c()
for(i in 1:ncol(csMEs)){
  me=csMEs[,i]
  mename=colnames(RepMEs)[i]
  groups=csMeta$week.4.remission
  namekey[namekey[,2]==mename,1]
  colorname=namekey[namekey[,2]==mename,1]
  df=data.frame("Eigengene"=me,"Response"=groups)
  pl=ggplot(df,aes(x=Response,y=Eigengene,fill=colorname))+geom_boxplot()+geom_quasirandom()+
    ylab(paste0(mename))+scale_fill_manual(values=colorname)+guides(fill=F)
  csplots[[mename]]=pl
  test=wilcox.test(me~groups)
  ps=c(ps,test$p.value)
}

adjp=p.adjust(ps,method="fdr")
siglist=csplots[adjp<0.05]

cat("**Wilcox results responders vs. non-responders corticosteroids (All)**")
psdf=data.frame("Module"=colnames(csMEs),"Wilcox_p"=ps,"Wilcox_FDR_p"=adjp)
psdf=psdf[order(adjp),]
kable(psdf)

write.csv(psdf,here("Data","Generated","Haberman_cs_repsonse_wilcox_module_eigengenes.csv"))

c("**Boxplots corticosteroid significatnly different (FDR p<0.05)**")
plot_grid(plotlist = siglist[psdf$Module[psdf$Wilcox_FDR_p<0.05]], ncol=3)
ggsave(here("Plots","Haberman_corticosteroids_signif_diff_by_response_boxplots.pdf"),width=10,height=15)

```

