---
title: "Additional Analyses"
output:
  html_document: default
  pdf_document: default
editor_options: 
  chunk_output_type: inline
---

Some additional analyses to address reviewer comments.

```{r echo=FALSE}
#library here lets things be relative to this repo
library(here)
#load BioFunk helper script
source(here("Scripts","BioFunk.R"))

#load other libraries
library(ggpubr)
library(tidyverse)
library(caret)
library(pROC)
library(edgeR)
library(ggvenn)
library(xCell)

#load WGCNA
suppressPackageStartupMessages(library(WGCNA))
options(stringsAsFactors = FALSE)
allowWGCNAThreads(nThreads = 20)
#set seed to make reproducible
set.seed(666)

theme_set(theme_cowplot())
```

# ROC Curves for predicting repsonse for M4 and M5 eigengenes

Carry out ROC curves at the eigengene level for M4 and M5.

## aTNF

```{r atnfroc, fig.height=4, fig.width=5}
#load the eigengenes for the aTNF study
atnfeig <- read.csv(here("Data","Generated","Arijs_2009_pre_treat_MEs.csv")) %>% select(c("M4","M5","Response"))
atnfeig$Response[atnfeig$Response=="Non-Responder"]="Non.Responder"
#run caret to find predictive ability for each module
control <- trainControl(method = "repeatedcv", number = 5, repeats = 10, summaryFunction = twoClassSummary, classProbs = TRUE)
set.seed(123)
reslist <- list()
for(i in 1:(ncol(atnfeig)-1)){
df <- data.frame(Module=atnfeig[,i],Response=atnfeig$Response)
fit <- train(Response ~ Module, data = df, method = "glm", 
             family = "binomial", trControl = control)
r <- fit$results$ROC
p <- predict(fit, df, type="prob")
rp <- roc(df$Response, p$Responder)
reslist[[i]] <- rp
}

plot(reslist[[1]],col="paleturquoise",main="Pre-treatment M4 eigengene\naTNF Response (Arijs 2009)")
text(x = 1, paste0("AUC:", round(reslist[[1]]$auc,2)))



plot(reslist[[2]],col="orange",main="Pre-treatment M5 eigengene\naTNF Response (Arijs 2009)")
text(x = 1, paste0("AUC:", round(reslist[[2]]$auc,2)))

```

## Corticosteroids

```{r cortroc, fig.height=4, fig.width=5}
#load the eigengenes for the cort study
corteig <- read.csv(here("Data","Generated","Haberman_pre_treat_corticosteroid_MEs.csv")) %>% select(c("M4","M5","Response"))
corteig$Response[corteig$Response=="Non-Responder"]="Non.Responder"

reslist <- list()
for(i in 1:(ncol(corteig)-1)){
df <- data.frame(Module=corteig[,i],Response=corteig$Response)
fit <- train(Response ~ Module, data = df, method = "glm", 
             family = "binomial", trControl = control)
r <- fit$results$ROC
p <- predict(fit, df, type="prob")
rp <- roc(df$Response, p$Responder)
reslist[[i]] <- rp
}

plot(reslist[[1]],col="paleturquoise",main="Pre-treatment M4 eigengene\nCorticosteroid Response (Haberman 2019)")
text(x = 1, paste0("AUC:", round(reslist[[1]]$auc,3)))

plot(reslist[[2]],col="orange",main="Pre-treatment M5 eigengene\nCorticosteroid Response (Haberman 2019)")
text(x = 1, paste0("AUC:", round(reslist[[2]]$auc,3)))

```

# DE Genes Responders vs HC as predictors of response 

Do genes that differ in reponsive disease also predict response. i.e. Are the genes that are different in responders compared to healthy controls the same as those that differ between responders and non-responders, just to a different extent.

```{r, fig.width=8, fig.height=5}
#load in the gene counts
genes <- read.table(here("Data","Input","Arijs_2009_cleaned_expr_dat.tsv.gz"),sep="\t",header=T)
meta <- read.table(here("Data","Input","Arijs_2009_cleaned_meta_table.tsv.gz"),sep="\t",header=T)

#do responders pre-treatment vs hc
hcm <- meta %>% filter(Control==1|PostTreatment==0) %>% filter(is.na(.$aTNF_Response)|aTNF_Response==1) %>% filter(Tissue==1)
hgen <- genes[,hcm$GID]

#model matrix
mm <- model.matrix(~0+Control+IBD_Combined,data = hcm)
fit <- lmFit(hgen,mm)
cont.matrix <- makeContrasts(IBDvsHC=IBD_Combined-Control, levels=mm)
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)
hcres <- topTable(fit2, adjust="BH", number = Inf)

top50 <- hcres %>% rownames_to_column() %>% arrange(desc(logFC)) %>% slice(1:50)

#do responders vs non-responders
respm <- meta %>% filter(Control==0&PostTreatment==0) %>% filter(Tissue==1)
respm$Responder <- ifelse(respm$aTNF_Response==1,1,0)
respm$NonResponder <- ifelse(respm$aTNF_Response==1,0,1)
resgen <- genes[,respm$GID]

#model matrix
rmm <- model.matrix(~0+Responder+NonResponder,data = respm)
rfit <- lmFit(resgen,rmm)
rcont.matrix <- makeContrasts(NonRespvsResp=NonResponder-Responder, levels=rmm)
rfit2 <- contrasts.fit(rfit, rcont.matrix)
rfit2 <- eBayes(rfit2)
respres <- topTable(rfit2, adjust="BH", number = Inf)


#get module list
modlist <- read_delim(here("Data/Generated/discovery_WGCNAgeneInfo.csv"),delim=",")

#merge results
rmatched <- respres[rownames(hcres),]
colnames(rmatched) <- paste0("resp_",colnames(rmatched))
merged <- cbind(rmatched,hcres) %>% data.frame()
rownames(merged) <- rownames(rmatched)
merged$Module <- modlist$moduleColor[match(rownames(merged),modlist$geneid)]
merged <- merged %>% filter(.$resp_adj.P.Val<0.05|.$adj.P.Val<0.05)
merged$Color <- ifelse(merged$Module%in%c("M4","M5"), merged$Module, "Other")
merged$T50 <- ifelse(merged$logFC>=min(top50$logFC),"Top50",NA)

scatter <- ggplot(merged,aes(x=resp_logFC,y=logFC,color=Color))+geom_point(aes(color=T50),size=2)+geom_point()+scale_color_manual(values=c("orange","paleturquoise","grey","red"))+xlab("LogFC Responders vs Non-Responders")+ylab("LogFC IBD Responders vs HC")

m4list <- which(merged$Module=="M4")
m5list <- which(merged$Module=="M5")
t50list <- which(merged$T50=="Top50")

ven <- ggvenn(list(M4=m4list,M5=m5list,Top50=t50list),fill_color = c("paleturquoise","orange","red"), show_percentage = F)

plot_grid(scatter,ven,nrow=1,rel_widths = c(0.7,0.3))

```

# xCell on prospective anti-TNF cohort

```{r xcell, fig.width=5, fig.height=4}
#give gene symbols#load in the gene counts
genes <- read.table(here("Data","Input","Arijs_2009_cleaned_expr_dat.tsv.gz"),sep="\t",header=T)
meta <- read.table(here("Data","Input","Arijs_2009_cleaned_meta_table.tsv.gz"),sep="\t",header=T)

#gene gene symbols
genes <- read.table(here("Data","Input","Arijs_2009_cleaned_expr_dat.tsv.gz"),sep="\t",header=T)
meta <- read.table(here("Data","Input","Arijs_2009_cleaned_meta_table.tsv.gz"),sep="\t",header=T)
genes$Symbol <- modlist$geneName[match(rownames(genes),modlist$geneid)]
genes <- genes %>% filter(!is.na(Symbol)) %>% group_by(Symbol) %>% summarise_all(sum)
gdf <- genes %>% data.frame()
rownames(gdf) <- gdf$Symbol
gdf <- gdf %>% select(-1)

#run xcell
xc <- xCellAnalysis(gdf)
neut <- xc["Neutrophils",]

meta$Neutrophils <- neut

pret <- meta %>% filter(Control==0,PostTreatment==0)
pret$aTNF_Response <- ifelse(pret$aTNF_Response==1,"Responder","Non-Responder")

pret %>% ggplot(aes(x=as.factor(aTNF_Response),y=Neutrophils))+geom_boxplot()+geom_point()+xlab("anti-TNF Response (Arijs 2009)")+
  ylab("xCell Predicted Neutrophil Score\n(Pre-treatment)")+stat_compare_means(label.x.npc = 0.9, comparisons = list(c(1,2)), method = "t.test")
```

