---
title: "Replication of WGCNA Modules in Arijs 2009 data"
output: html_document
editor_options: 
chunk_output_type: console
---

# Load in the replication data

```{r echo=FALSE}
#load BioFunk
source("~/devel/Useful_Scripts/BioFunk.R")
#load WGCNA
suppressPackageStartupMessages(library(WGCNA))
allowWGCNAThreads()
options(stringsAsFactors = FALSE)
suppressPackageStartupMessages(library("GEOquery"))
suppressPackageStartupMessages(library("affy"))
suppressPackageStartupMessages(library(ggbeeswarm))
suppressPackageStartupMessages(library(ggpubr))
library(here)
library(knitr)
library(cowplot)
set.seed(666)
theme_set(theme_cowplot())

#### the below commands were originally used to generate the normalised data ####

#getGEOSuppFiles("GSE16879")
#untar("GSE16879_RAW.tar",exdir="data")
#cels=list.files("data/",pattern="CEL")
#setwd("data/")
#raw.data = ReadAffy(verbose = FALSE, filenames = cels)
#data.rma.norm = rma(raw.data)
#rma = exprs(data.rma.norm)
#write.table(rma, file = "normalised_array_exp.tsv", quote = FALSE, sep = "\t")

#read in the normalised data
#expdat=read.table("normalised_array_exp.tsv",header=T,sep="\t")
#mapped=t(expdat)
#convert the probes names to gene ids
#library(hgu133plus2.db)
#mapping=data.frame(hgu133plus2ENSEMBL)
#colnames(mapped)=mapping$ensembl_id[match(colnames(mapped),mapping$probe_id)]
#mapped=mapped[,-which(is.na(colnames(mapped)))]
#mapped=mergDupCol(mapped,TRUE)
#rownames(mapped)=gsub(".CEL.gz","",rownames(mapped))
#clean_expr_dat=t(mapped)
#write.table(clean_expr_dat,"Arjis_2009_cleaned_expr_dat.tsv",sep="\t")

repdat=read.table(here("Data","Input","Arijs_2009_cleaned_expr_dat.tsv.gz"),sep="\t")

#### the code below was originally used to generate the metadata table ####

#load the metadata downloaded from the GEO website
#geodat=getGEO(filename="GSE16879_series_matrix.txt.gz")
#geometa=pData(geodat)
#metasmall=geometa[,c(1,2,36,37,38,39)]
#pt=ifelse(metasmall$`before or after first infliximab treatment:ch1`=="After first infliximab treatment",1,0)
#pt[metasmall$`before or after first infliximab treatment:ch1`=="Not applicable"]=NA
#re=ifelse(metasmall$`response to infliximab:ch1`=="Yes",1,0)
#re[metasmall$`response to infliximab:ch1`=="Not applicable"]=NA
#meta=data.frame(
#  GID=metasmall$geo_accession,
#  Sample_Name=metasmall$title,
#  Individual=unlist(lapply(strsplit(metasmall$title,"_"),"[[",1)),
#  PostTreatment=pt,
#  Control=ifelse(metasmall$`disease:ch1`=="Control",1,0),
#  UC=ifelse(metasmall$`disease:ch1`=="UC",1,0),
#  CD=ifelse(metasmall$`disease:ch1`=="CD",1,0),
#  IBD_Combined=ifelse(metasmall$`disease:ch1`=="CD"|metasmall$`disease:ch1`=="UC",1,0),
#  aTNF_Response=re,
#  Tissue=ifelse(metasmall$`tissue:ch1`=="Colon",1,0)
#)
#write.table(meta,"Arijs_2009_cleaned_meta_table.tsv",sep="\t")

metadat=read.table(here("Data","Input","Arijs_2009_cleaned_meta_table.tsv.gz"),sep="\t")
metadat=data.frame(lapply(metadat,factor))
metadat=rowNamer(metadat)
metadat=metadat[match(rownames(metadat),colnames(repdat)),]
metadat=metadat[,-which(colnames(metadat)=="Sample_Name")]
```

Microarray data was downloaded from the Arijs et al. 2009 PLoS One paper as raw image data from (GEO:GSE16879).
This contained **`r nrow(metadat)`** samples and **`r nrow(repdat)`** genes.
Microarray data was normalised using the robust multi-array average (RMA) measure.

# Map to discovery cohort modules

```{r, echo=FALSE}
#our module data
moddefs=read.csv(here("Data","Generated","discovery_WGCNAgeneInfo.csv"))
modcors=read.csv(here("Data","Generated","discovery_WGCNAmodInfo.csv"))
moddefs=moddefs[-which(moddefs$moduleColor=="unassigned"),]

#list of modules significantly associated with NANCY score
signancy=modcors[modcors$NANCYscore_Pvalue_FDR_Pvalue<0.05,]

#match replication data and module defs
matchdef=moddefs[moddefs$geneid%in%rownames(repdat),]
matchrep=repdat[match(matchdef$geneid,rownames(repdat)),]

#calculate the module coverages in the replication data set (gene % observed)
modcov=data.frame(Original=table(moddefs$moduleColor),Replication=table(matchdef$moduleColor))
modcov=modcov[,-3]
colnames(modcov)=c("Module","Original Size (No. Genes)","Replication Size (No. Genes)")
modcov$Percentage=(modcov[,3]/modcov[,2])*100
modcov$NANCYsig=ifelse(modcov$Module%in%signancy$X,1,0)
colnames(modcov)[ncol(modcov)]="FDR Sig. with NANCY Score in discovery cohort."
modcov=modcov[order(modcov$Percentage,decreasing = T),]

```

In our unbiased analysis we had **`r nrow(moddefs)`** genes assigned to a module (excluding the genes assigned to grey).
From these, **`r nrow(matchdef)`** were also in the replication data set (**`r round((nrow(matchdef)/nrow(moddefs))*100)`%**).

## Module overlaps

`r kable(modcov,format="markdown",row.names=F)`

# Module eigengene differences in pre-treatment samples by treatment response

```{r echo=FALSE, fig.width=10, results="asis"}
#generate the module eigengenes
RepMEList = moduleEigengenes(t(matchrep), colors = matchdef$moduleColor, excludeGrey = TRUE,)
RepMEs = RepMEList$eigengenes
colnames(RepMEs)=gsub("ME","",colnames(RepMEs))

#correlation between MEs in this dataset
betMEcors = cor(RepMEs[which(metadat$PostTreatment==0),])
nSamples=nrow(RepMEs[which(metadat$PostTreatment==0),])
betMEps = corPvalueStudent(betMEcors, nSamples)
betMEpadj=matrix(p.adjust(betMEps,method="fdr"),ncol=38,byrow = F)
colnames(betMEpadj)=colnames(betMEps)
rownames(betMEpadj)=rownames(betMEps)
betMEpadj=ifelse(betMEpadj<0.05,1,0)
colorer=colorRampPalette(rev(brewer.pal(11,"RdBu")))
simpHeat(betMEcors,betMEpadj,colorer(11))
ggsave(here("Plots","Arijis_2009_pre_treat_correlation_between_MEs.pdf"),width = 8,height = 6)

#remove non NANCY associated MEs
RepMEs = RepMEs[,colnames(RepMEs)%in%signancy$X]

```

## Pre-treatment differences by response 

```{r, fig.height=15, fig.width=10,results="asis",echo=F}
#get the color mapping from the name
namekey=read.csv(here("Data","Input","module_number_key.csv.gz"),header=F)

preMEs=RepMEs[which(metadat$PostTreatment==0),]
preMeta=metadat[which(metadat$PostTreatment==0),]

preMeta$aTNF_Response=as.character(preMeta$aTNF_Response)
preMeta$aTNF_Response[preMeta$aTNF_Response==1]="Yes"
preMeta$aTNF_Response[preMeta$aTNF_Response==0]="No"
preMeta$aTNF_Response=factor(preMeta$aTNF_Response)

#write out the pretreatment MEs
writeMEs=preMEs
writeMEs$Response=ifelse(preMeta$aTNF_Response=="Yes","Responder","Non-Responder")
write.csv(writeMEs,here("Data","Generated","Arijs_2009_pre_treat_MEs.csv"))

preplots=list()
ps=c()
for(i in 1:ncol(preMEs)){
  me=preMEs[,i]
  mename=colnames(preMEs)[i]
  groups=preMeta$aTNF_Response
  colorname=namekey[namekey[,2]==mename,1]
  df=data.frame("Eigengene"=me,"Response"=groups)
  pl=ggplot(df,aes(x=Response,y=Eigengene,fill=colorname))+geom_boxplot()+geom_quasirandom()+
    ylab(paste0(mename))+scale_fill_manual(values=colorname)+guides(fill=F)
  preplots[[mename]]=pl
  test=wilcox.test(me~groups)
  ps=c(ps,test$p.value)
}

adjp=p.adjust(ps,method="fdr")
siglist=preplots[adjp<0.05]

cat("**Wilcox results responders vs. non-responders pre-treatment**")
psdf=data.frame("Module"=colnames(preMEs),"Wilcox_p"=ps,"Wilcox_FDR_p"=adjp)
psdf=psdf[order(adjp),]
kable(psdf)

write.csv(psdf,here("Data","Generated","Arijs_2009_pretreatment_wilcox_eigengenes.csv"))

c("**Boxplots for FDR significant results**")
plot_grid(plotlist = siglist[psdf$Module[psdf$Wilcox_FDR_p<0.05]], ncol=4)
ggsave(here("Plots","Arijs_2009_reponder_pre_treatment_boxplots.pdf"),width=13.33,height=30)
```

# Effect size information between responders and non-responders for M4 & M5 for meta analysis

```{r effsize_M4_M5, include=T}

M4=preMEs$M4
M5=preMEs$M5
response=preMeta$aTNF_Response

ne=length(which(response=="No"))
nc=length(which(response=="Yes"))

me4=mean(M4[response=="No"])
mc4=mean(M4[response=="Yes"])

sde4=sd(M4[response=="No"])
sdc4=sd(M4[response=="Yes"])

me5=mean(M5[response=="No"])
mc5=mean(M5[response=="Yes"])

sde5=sd(M5[response=="No"])
sdc5=sd(M5[response=="Yes"])

efftab=rbind(c(ne,me4,sde4,nc,mc4,sdc4),c(ne,me5,sde5,nc,mc5,sdc5))
colnames(efftab)=c("N Non","Mean Non","SD Non","N Resp","Mean Resp","SD Resp")
rownames(efftab)=c("M4","M5")

kable(efftab)

```